---
title: "R Notebook"
output: html_notebook
---

`PMGs` procedure is illustrated on a cirrhosis study data. Data is obtained from the <https://github.com/knights-lab/MLRepo>.

Load Cirrhosis Data

```{r}
library(readr)
tax_otu <- read.table('data/processed_taxatable.txt', sep="\t",header = TRUE) 
tax<-tax_otu[,1:7]
otu<-tax_otu[,-(1:7)]
labels <- read.table('data/task-healthy-cirrhosis.txt', sep="\t",header = FALSE) 
```

A Phyloseq object created. Data is filtered and 0's are replaced. TAX table and Sampledata are organized.

```{r}
library(phyloseq)
OTU<-otu_table(otu, taxa_are_rows=TRUE) # taxas x samples
TAX<-tax_table(as.matrix(tax))
SAMPLEDATA<-sample_data(as.data.frame(labels))
taxa_names(OTU)<-taxa_names(TAX)
sample_names(OTU)<-sample_names(SAMPLEDATA)
cir<- phyloseq(OTU,TAX,SAMPLEDATA)
#Taxa that were not seen with more than 20 counts in at least 30% of samples are filtered.
cir2 <-  filter_taxa(cir, function(x) sum(x > 20) > (0.3*length(x)), TRUE)
OTU<-t(as.data.frame(otu_table(cir2)))
library(zCompositions)
otu.no0<-cmultRepl(OTU)
TAX<-as.data.frame(tax_table(cir))
TAX$otuid<-rownames(TAX)
SAMPLEDATA$Label<-ifelse(grepl("Cirrhosis", SAMPLEDATA$V2), 1, 0)
names(SAMPLEDATA)<-c("sampleID","Status","Label")
```

Number of `PMGs` depends on the number of principal balances (PBs) selected. Before `PMGs` construction, the analyst should decide the minimum number of PBs to include into `PMGs` construction. 

Looking at variance plot of PBs, the most informative PBs are first two and choosing the first two principal balance results 3 dense OTU groups. The aim is to get as many groups as possible that explain the most of the total variance in the dataset with a manageable number of OTUs in it. For the demo data, minimum number is 10. 

```{r}
X<-otu.no0
V<- mPBclustvar(X) #
coord<-milr(X,V)
plot(diag(var(coord)),xlab="number of coordinates",ylab="Explained Variance")
minimumNumOfGroups<-11
numberofgroup<-findOptimalNumOfGroups(otu.no0,minimumNumOfGroups)
```

Optimal Number of groups is calculated by the method findOptimalNumOfGroups(otu,min).

```{r}
minimumNumOfGroups<-11
numberofgroup<-findOptimalNumOfGroups(otu.no0,minimumNumOfGroups)
```

Once optimal number is decided, then it is easy to construct PMGs on otu table.
The otuput is PMG table and and OGT(OtuID|Group|Taxa) dataframe.

```{r}
PMGs<-createPMGs(otu.no0,numberofgroup)  
head(OGT)
```

Draw a CODA Dendrogram on PMGs

```{r}
PMGs_labels<-cbind(PMGs,SAMPLEDATA$Status)
names(PMGs_labels)[ncol(PMGs_labels)]<-"label"
PMGs_labels$label<-as.factor(PMGs_labels$label)
W<- mPBclustvar(PMGs)
library(compositions)
CoDaDendrogram(X=acomp(PMGs),V=W,type="lines",range=c(-10,10))
CoDaDendrogram(X=acomp(PMGs[PMGs_labels$label=="Cirrhosis",]), col="red",add=TRUE,V=E,type="lines",range=c(-7,7))
CoDaDendrogram(X=acomp(PMGs[PMGs_labels$label=="Healthy",]), col="green",add=TRUE,V=E,type="lines",range=c(-7,7))
```

Find Compositional Biomarkers via Distal PMG Balances

```{r}
library(caret)
library(balance)

fit.control <- trainControl(method = "repeatedcv", number = 10, repeats = 5,
                            summaryFunction = twoClassSummary, classProbs = TRUE, allowParallel = F, savePredictions = T)

sbp <- sbp.fromADBA(PMGs, SAMPLEDATA$Status) # get discriminant balances
sbp <- sbp.subset(sbp) # get distal balances only

compBiomarkers <- as.data.frame(balance.fromSBP(x=PMGs, y = sbp)) 
compBiomarkers.labeled <- addLabel(data.distalBal.pmg,SAMPLEDATA$Status)


set.seed(123)
fit <- train(label ~ ., data = compBiomarkers.labeled, method = "glm", 
                          family = "binomial", trControl = fit.control)
fit$results
```

